---
author: "Hera Fatma (231180006)"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
DATASET: Student Mental Health Survey (Source: Kaggle)


LOADING LIBRARIES 
```{r}
# install.packages("corrplot")
library(tidyverse)
library(ggplot2)
library(lsr)
library(psych)
library(dplyr)
library(corrplot)
library(reshape2)
library(viridis)
```
LOAD DATASET

```{r}
data_health <- read_csv("./../datasets/MentalHealthSurvey.csv")
```
**EXPLORATORY DATA ANALYSIS:**

*STATISTICAL ANALYSIS:*


Aim: To check how all the other factors effects Mental health indicators: depression, anxiety, isolation, future_insecurity.

```{r}
head(data_health)
dim(data_health)
sample(data_health) ## sample random rows
spec(data_health)  ## column specifications
str(data_health)
```

```{r}
summary(data_health)
```
```{r}
describe(data_health) ## categorical data
```


CHECKING FOR NULL AND DUPLICATE VALUES

```{r}
sum(is.na(data_health))
sum(duplicated(data_health))
```
Inference: no missing or duplicated value. Therefore, no data cleaning required.


SEPERATING COLUMNS BASED ON DATA TYPES

1. TO SELECT ALL THE COLUMNS WHICH ARE NUMERIC

```{r}
data_num <- select_if(data_health, is.numeric)
data_num
```
2. TO SELECT ALL THE COLUMNS WHICH ARE CHARACTER

```{r}
data_char <- select_if(data_health, is.character)
data_char
```


*CENTRAL TENDENCIES*:


Mean for all the numerical columns

```{r}
mean_cols <- colMeans(
select (data_num, age, study_satisfaction, academic_workload, academic_pressure, financial_concerns, social_relationships, depression, anxiety, isolation, future_insecurity))
mean_cols
```


Standard Deviation for all the numerical columns

```{r}
# age
sd_age <- sd(data_health$age,2)
sd_age

# all num_cols
sd_cols <- apply(data_health[, 12:20], 2, sd)
sd_cols
```


SKEWNESS

Note:
Skewness values and interpretation

A rule of thumb states that: 

Symmetric: Values between -0.5 to 0.5. 
Moderated Skewed data: Values between -1 and -0.5 or between 0.5 and 1. 
Highly Skewed data: Values less than -1 or greater than 1.

```{r}
skew(data_num)
```

KURTSIS

Main three types of kurtosis.

Mesokurtic: This is the normal distribution
Leptokurtic: This distribution has fatter tails and a sharper peak. The kurtosis is “positive” with a value greater than 3
Platykurtic: The distribution has a lower and wider peak and thinner tails. The kurtosis is “negative” with a value greater than 3

*Note: that we define the excess kurtosis as kurtosis minus 3

```{r}
describe(data_num)
```

```{r}
describeBy( x=data_health, group=data_health$gender)
```

```{r}
# describeBy( x=data_health, group=data_health$depression)
```

```{r}
# describeBy( x=data_health, group=data_health$anxiety)
```

```{r}
# describeBy( x=data_health, group=data_health$isolation)
```

```{r}
by(data=data_health, INDICES=data_health$gender, FUN=summary)
```
Inference: 

Females are more prone to an imbalance in mental health.

```{r}
# anxiety, depression, acadmic_workload

aggregate(anxiety ~ depression + academic_workload, 
           data = data_health,
           FUN = mean)
```
Inference:

High workload and depression score leds to more anxiety.


```{r}
aggregate(anxiety ~ depression + age, 
           data = data_health,
           FUN = mean)
```
Inference:

more age + high depression score = more anxiety levels


```{r}
aggregate(anxiety ~ academic_workload + academic_pressure, 
           data = data_health,
           FUN = mean)
```
Inference:

High academic workload is leds to high anxiety.



*DATA DISTRUBUTION* 

```{r}
# age
ggplot(data_health, aes(x = age)) + geom_histogram(bins = 10, color = "black", fill = "dodgerblue3")

# depression
ggplot(data_health, aes(x = depression)) + geom_histogram(bins = 5, color = "black", fill = "hotpink2")
```


*DEMOGRAPHIC DISTRIBUTIONS (PLOTS)*

1. BAR PLOTS

```{r}
# Distribution of Categorical variables

cat_vars <- c("gender", "university", "degree_level", "degree_major", "academic_year", "cgpa", "residential_status", "campus_discrimination", "sports_engagement", "average_sleep" )

for (var in cat_vars) {
  print(ggplot(data_health, aes_string(x = var)) +
    geom_bar(color = "black", fill = "dodgerblue3") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste("Distribution of", var))) +
    theme_classic()
}
```

Inference:

1. No.of male >> no. of female.
2. PU university has the maximum admission.
3. Undergrads >> Postgrads.
4. Max. students in Data science and computer science.
5. More number of students in 1st and 3rd year.
6. Max with cgpa between 3.0 to 4.0
7. Majority stays off- campus.
8. Less people faced campus discrimination.
9. Majority are not involved in any sports.
10. Majority has average sleep between 4-6 hrs.



2. Distribution of numerical variables

```{r}
# Distribution of numerical variables

num_vars <- c("study_satisfaction", "academic_workload", "academic_pressure", "financial_concerns", "social_relationships", "depression", "anxiety", "isolation", "future_insecurity")

for (var in num_vars) {
  print(ggplot(data_health, aes_string(x = var)) +
    geom_histogram(binwidth = 1, color = "black", fill = "hotpink3") +
    labs(title = paste("Distribution of", var)))
}
```

3. BOXPLOT

```{r}
combined_data <- data_health %>%
  pivot_longer(cols = c(study_satisfaction, academic_workload, academic_pressure, financial_concerns,  social_relationships, depression, anxiety, isolation, future_insecurity),
              names_to = "Variable",
              values_to = "Value")


ggplot(combined_data, aes(x = Variable, y = Value, fill = Variable)) +
geom_boxplot() +
labs(title = "Sleep Dataset") +
theme_classic() +
theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_fill_manual(values = c("cornflowerblue", "maroon", "cornflowerblue", "maroon", "cornflowerblue", "maroon", "cornflowerblue", "maroon", "cornflowerblue", "maroon", "cornflowerblue"))
```

Inference: No outliers


4. Pie-chart (Gender)

```{r}
# Pie chart: Gender distribution

gender_counts <- data_health %>% count(gender)
ggplot(gender_counts, aes(x = "", y = n, fill = gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Gender Distribution") +
  scale_fill_brewer(palette = "Set6")
```

3. NO. OF MALE AND FEMALE STUDENTS BASED ON UNIVERSITY, DEGREE_LEVEL AND DEGREE_MAJOR

```{r}
ggplot(data_health, aes(gender, degree_major, fill = degree_level)) + geom_bar(stat = "identity", position = "dodge") + theme_classic()

ggplot(data_health, aes(gender, university, fill = degree_level)) + geom_bar(stat = "identity", position = "dodge") + theme_classic()

ggplot(data_health, aes(university, gender, fill = degree_level)) + geom_bar(stat = "identity", position = "dodge") + theme_classic()

```
```{r}
# Filtering male and female data 

no_male <- filter(data_health, gender == "Male")
dim(no_male)
no_female <- filter(data_health, gender == "Female")
dim(no_female)
```

Inference:

Branch and degree-wise distribution of students.

1. The dataset appears to have more male students than female students.
2. No female student in postgraduate.
3. Most students are undergraduates, with a mix of Data Science, Computer Science, IT and SWE majors.


To check number of PG Females

```{r}
## To check number of PG Females

females <- filter(data_health, gender == 'Female')
females
pg_females <- sum(filter(females, degree_level == 'Postgraduate'))
pg_females
dim(females)
dim(pg_females)
```
Inference:

1. A total of 24 female students in overall program.
2. Conformation --> no female pg student.


4. Barplots

```{r}
#barplot(as.matrix(data_num$isolation), main="Multiple Bar Plots", ylab="Count", beside=TRUE)
```


5. Violin plot (Depression scores by gender)

```{r}
ggplot(data_health, aes(x = gender, y = depression, fill = gender)) +
  geom_violin() +
  geom_boxplot(width = 0.1, fill = "white") +
  theme_classic() +
  labs(title = "Depression Scores by Gender", x = "Gender", y = "Depression Score") +
  scale_fill_brewer(palette = "Pastel1")
```
Inference:

Female have a higher depression score.


6. Scatter plot (Relationship between academic pressure and depression)

```{r}
ggplot(data_health, aes(x = academic_pressure, y = depression)) + geom_point(aes(color = gender), alpha = 0.6) + geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(title = "Academic Pressure vs Depression", x = "Academic Pressure", y = "Depression Score") +
  scale_color_brewer(palette = "Set1")
```
Inference:

1. Linearly correlated.
2. As the academic pressure increases, depression score also increases.



7. Stacked bar plot (CGPA distribution by degree major)

```{r}
ggplot(data_health, aes(x = degree_major, fill = cgpa)) +
  geom_bar(position = "fill") +
  theme_classic() +
  labs(title = "CGPA Distribution by Degree Major", x = "Degree Major", y = "Proportion") +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

8. Grouped bar plot  (Average sleep duration by sports engagement)

```{r}
ggplot(data_health, aes(x = sports_engagement, fill = average_sleep)) +
  geom_bar(position = "dodge") +
  theme_classic() +
  labs(title = "Average Sleep Duration by Sports Engagement", x = "Sports Engagement", y = "Count") +
  scale_fill_brewer(palette = "YlOrRd") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Inference:

1. People with no sports engagements have an average sleep of 4-6 hrs.
2. More sports engagement effects the optimal sleep duration (7-8 hrs), negatively.



9. Density plot (Distribution of anxiety scores by gender)

```{r}
ggplot(data_health, aes(x = anxiety, fill = gender)) +
  geom_density(alpha = 0.7) +
  theme_classic() +
  labs(title = "Distribution of Anxiety Scores by Gender", x = "Anxiety Score", y = "Density") +
  scale_fill_brewer(palette = "Set2")
```
Inference:

Females have a higher distribution for anxiety score.


10. Faceted histogram (Mental health indicators: depression, anxiety, isolation, future_insecurity)

```{r}
mental_health_long <- data_health %>%
  select(depression, anxiety, isolation, future_insecurity) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(mental_health_long, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 1, alpha = 0.7) +
  facet_wrap(~ variable, scales = "free_y") +
  theme_classic() +
  labs(title = "Distribution of Mental Health Indicators", x = "Score", y = "Count") +
  scale_fill_brewer(palette = "Set1")
```


11. CORRELATINAL PLOT BETWEEN NUMERICAL VARIABLES

```{r}
#cor_matrix <- cor(data_health %>% select(study_satisfaction, academic_workload, academic_pressure, 
                                                #financial_concerns, social_relationships, depression, 
                                                #anxiety, isolation, future_insecurity))


cor_matrix <- cor(data_health[num_vars])
corrplot(cor_matrix, method = "circle")
cor_matrix
```

Heat Map

```{r}
# Heatmap

melted_cor_matrix <- melt(cor_matrix)
ggplot(melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", high = "blue", mid = "white", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap of Mental Health Variables")
```

Inference:

High correlation:

    1. Anxiety  and depression
    2. Isolation and depression
    3. Anxiety and isolation
    
Low correlation:

    1. academic pressure and academic workload
    2. depression, anxiety, future insecurity with academic workload
    3. future insecurity with depression, anxiety and isolation
    
    
No correlations:

    1. Social relationships
    2. Study satisfaction
    
    
    
CONCLUSIONS:

1. Demographic Distribution:

The dataset appears to have more male students than female students.
Most students are undergraduates, with a mix of Data Science and Computer Science majors.

2. Academic Performance:

Most students have a CGPA between 3.0-4.0
There's a relatively even distribution across academic years (1st to 3rd year).


3. Mental Health Indicators:

The distributions of depression, anxiety, and isolation scores show variation, with some students reporting high levels.
There appears to be a correlation between depression, anxiety and isolation scores.


4. Academic Pressure and Workload:

Many students report high levels of academic pressure and workload.
High academic workload is leds to high anxiety.


5. Sleep Patterns:

Most students report sleeping 4-6 hours per night, which is less than the recommended amount for young adults.


6. Sports Engagement:

A significant portion of students report no sports engagement, which could impact overall well-being.





