---
author: "Hera Fatma (231180006)"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1.2. Addition and Subtraction 
```{r}
2+2
3-2
```
Error
```{r}
# 3 - 
  
  # ERROR: Error: Incomplete expression: 3
```

Division and Multiplication 
```{r}
3 / 2
3 * 2
```
Power
```{r}
2 ^ 2
2 ^ 3
```
1.3. OPERATIONS
```{r}
(2+3) * 3
2 + (3 * 3)
```
FUNCTIONS
```{r}
sqrt(4)
abs(-2)
abs(2)
```
```{r}
# sqrt() # function will display error without any argument

# ERROR: Error in sqrt() : 0 arguments passed to 'sqrt' which requires 1
```

1.4. ASSIGNING VARIABLES
```{r}
x <- 2 * 3
y = 2 -1
x
y 
# X # R is case sensitive. ERROR: Error: object 'X' not found
```
To retrieve a list of objects in your current working directory
```{r}
ls()
```
1.5. NUMERICAL VECTORS
```{r}
x <- c(2.3, 1, 5)
x # overwrites the value of x
```
To check the length of a vector
```{r}
length(x)
```
functions to access vector type
```{r}
mode(x)
class(x)
```
To create a sequence of vector using ':' function
```{r}
mynums <- 10:1
mynums
```
SUMMARY FUNCTIONS
```{r}
sum(mynums)
min(mynums)
max(mynums)
range(mynums) #min and max together
diff(range(mynums))
mean(mynums)
sd(mynums)
median(mynums)
```
If you use a function such as subtraction or division on a numeric vector, the function is repeated for all entries of the vector.
```{r}
mynums - 5 # subtract 5 from every number
mynums / 2 # divide every number by 2
```
1.6. INDEXING
```{r}
mynums[1]
mynums[2]
mynums[1:4]
mynums[-2] # will retrieve except 2nd position
```
```{r}
1:100
```
1.7. LOGICAL VECTORS

If u don't know the position of ur value of interest, check for values greater or less than.
```{r}
mynums > 3
mynums >= 3
```
```{r}
mynums < 4
mynums <= 4
mynums == 4
mynums != 4
```
NOTE: Result of logical operation is vector itself.

```{r}
mylog <- mynums >= 3
class(mylog)
```
NOTE: Logical vectors can be used for indexing.

```{r}
mynums[mylog]
```
```{r}
mynums [mynums >= 3]
```
1.8. CHARACTER VECTOR

```{r}
gender <- c('F', 'M', 'M', 'F', 'F', 'F')
gender
class(gender)
gender[2]   #indexing
gender[gender == 'F']
```
To get the index of a character

```{r}
which (gender == 'F')
```
Can't perform mathematical functions on character or logical vectors

```{r}
# mean(gender)

#ERROR: Warning: argument is not numeric or logical: returning NA[1] NA
```

1.9. FACTOR VECTORS 

NOTE: Factor has levels (i.e, different unique categories of the vector. same as the result of .unique() in python)
```{r}
gender <- as.factor(gender)
gender # output has no double quotes, unlike character vector.
levels(gender)
```
manipulating the vector

```{r}
gender[3] <- 'not_declared'
gender
```
```{r}
# To insert the values you first need to change the levels

levels(gender) <- c('M', 'F', 'not_declared')
gender[3] <- 'not_declared'
gender
```
1.10. DATA FRAMES

* 2-D object with rows and columns. Each columns contains a vector.
*Each argument of the function becomes a column.

```{r}
students <- c('Harry', "Sam", 'Abby')
df <- data.frame(students, score = c(50,77,89))
df
nrow(df)
ncol(df)
colnames(df)
df$score
mean(df$score)
str(df)
summary(df)
```

Indexing Data Frame

df[row , column]

```{r}
df[1,]
df[,2]
df[1:2,]
df[,1][2]  # first column 2nd element
```
To extract the row of the participant called sam using logical operand
```{r}
df[df$students == 'Sam',]
df[df$students == 'Sam',]$score
```

1.11. LOADING IN FILES

```{r}
getwd()
list.files()

# setwd()   ## To change working directory
```
Reading a file
```{r}
nettle <- read.csv("./../datasets/nettle_1999_climate.csv")

head(nettle)
```
```{r}
tail(nettle)
```
Overwriting contents of a data frame --> can be corrected. Solution: simply re-execute the above 3 blocks (from file loding).
```{r}
# nettle$Langs <- NA   ## overwrites Langs column as NA
nettle
```
Read .txt files (tab seperated)
```{r}
mydf <- read.table("./../datasets/sample1.txt", sep = '\t', header = TRUE)
```

1.12. PLOTTING

```{r}
hist(nettle$Langs)
```
```{r}
hist(nettle$Langs, col = 'salmon')
hist(nettle$Langs, col = "maroon")
```
```{r}
head(colors())
```
1.13. INSTALLING, LOADING AND CITING PACKAGES

a. install package --> install.packages('')
b. load package --> library()

```{r}
# install.packages("tidyverse")
```
```{r}
library(tidyverse)
```
citing packages
```{r}
citation('tidyverse')
packageVersion('tidyverse')
```
```{r}
R.Version()
```
1.14. SEEKING HELP

syntax: ?function name

```{r}
?seq()
```
If u forget a function name
```{r}
head(apropos('test'))
```










NOTES:

1. Type casting: syntax --> as.data_type()
2. In factor --> DATA POINTS are TOKENS of UNIQUE CATOGERY.
3. Result of logical operation is vector itself.
4. Logical vectors can be used for indexing